---
title: "Structural Divergence Analysis"
author: "María Laura Marcos, Julian Echave"
date: "25 de julio de 2016"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = 2, autodep = T)
```


```{r load_libraries}
library(ggplot2)
library(reshape2)
library(plyr)
library(quantreg)
library(bio3d)

setwd("C:/Users/Usuario/Desktop/VariabilidadEstructuralProteica")

source("OUT/out_subset_CA/multiplot.R")
source("OUT/out_subset_CA/my-functions.R")
```

```{r set_my_colors, eval = T}
myColors = c("blue", "darkgreen", "red", "red")
names(myColors) <- c("exp", "mut", "strong", "mut+sel")

setwd("C:/Users/Usuario/Desktop/VariabilidadEstructuralProteica/OUT")

```
# Introduction
We study the evolutionary divergence of proteins using the LFENM and compare with experimental data.

```{r set_run_variables}
# Set options
family = "globins"
R0 = 7.5
enm = "ANM"

# significance factors
f95 = 1.96
f99 = 2.6
f999 = 3.3
data.dir = paste("out_subset_CA", sep = "")
```


```{r set_file_names}
## Input and pre-processing of data
# Set up input filenames

reference.fname = file.path(data.dir, paste(family, "_ref.txt", sep = ""))
site.info.fname = file.path(data.dir, paste(family, "_out_m.da.csv", sep = ""))
protein_list.fname = file.path(data.dir, paste(family, "_list.txt", sep = ""))

dri2.exp.fname = file.path(data.dir, paste(family, "_R0_10_beta_strong.sel_K.analysis_Keff_out_m.exp.smooth.norm.dr.squarei.csv", sep = ""))
Pn.exp.fname = file.path(data.dir, paste(family,"_R0_10_beta_strong.sel_K.analysis_Keff_out_m.exp.Pn.csv", sep = ""))
energy.exp.fname = file.path(data.dir, paste(family, "_R0_10_beta_strong.sel_K.analysis_Keff_out_m.exp.va.csv", sep = ""))

dri2.strong.fname = file.path(data.dir, paste(family, "_R0_10_beta_strong.sel_K.analysis_Keff_out_m.theo.smooth.norm.dr.squarei.csv", sep = ""))
Pn.strong.fname = file.path(data.dir, paste(family, "_R0_10_beta_strong.sel_K.analysis_Keff_out_m.theo.Pn.csv", sep = ""))
energy.strong.fname = file.path(data.dir, paste(family, "_R0_10_beta_strong.sel_K.analysis_Keff_out_m.theo.va.csv", sep = ""))

dri2.mut.fname = file.path(data.dir,paste(family, "_R0_10_beta_no.sel_K.analysis_Keff_out_m.theo.smooth.norm.dr.squarei.csv", sep = ""))
Pn.mut.fname = file.path(data.dir, paste(family, "_R0_10_beta_no.sel_K.analysis_Keff_out_m.theo.Pn.csv", sep = ""))
energy.mut.fname = file.path(data.dir, paste(family, "_R0_10_beta_no.sel_K.analysis_Keff_out_m.theo.va.csv", sep = ""))

```

```{r set_reference_protein}
## which is the reference protein?
reference = as.character(read.table(reference.fname, header = F)[1, 1])
```

We use the `r enm` Elastic Network Model with `R0 =` `r R0` for the `r family` family. 
We use as reference for the ENM the protein with pdb code `r reference`.
The `r family` family consists of the following proteins:

```{r set_experimental_protein_names}
## Experimental protein names
protein.exp = read.table(protein_list.fname, header = F, stringsAsFactors = F)[, 1]
protein.exp = protein.exp[protein.exp != reference]
print(protein.exp)
```

[//]: Prepare site.info data frame

```{r set_up_site_info, eval = T}

## read site information of reference protein
site.info = read.csv(site.info.fname, header = T, sep = ";")

## add info to site.info
d = ddply(site.info, "site", .fun = function(x) data.frame("da" = min(x[, c(-1, -2)])))

## Define shells of active site neighborhood (first shell is from 2.5 to 7.5, second from 7.5 to 12.5, etc)
d$shell = cut(d$da, breaks = c(-.1, 2.5, 7.5, 12.5, 17.5, 22.5, 27.5, 32.5, 37.5, 42.5, 47.5, 52.5, max(d$da)))
levels(d$shell) = seq(length(levels(d$shell))) - 1 # rename shell levels
site.info = merge(site.info, d, by = "site")

## Read reference pdb and calculate resno, AA, and CN
reference.pdb.fname = file.path(data.dir,paste(reference, ".pdb", sep = ""))
ref.pdb = read.pdb(reference.pdb.fname)
inds = atom.select(ref.pdb, elety = "CA")
ref.pdb.ca = ref.pdb$atom[inds$atom,]
cmap.ca = cmap(ref.pdb$xyz[inds$xyz], dcut = R0, scut = 0, mask.lower = F)
CN = rowSums(cmap.ca, na.rm = T) - 1 # diag(cmap.ca) returned by cmap is 1!
AA = ref.pdb.ca$resid
resno = ref.pdb.ca$resno
pdb.info = data.frame(resno, AA, CN)
names(pdb.info) = c("resno", "AA", paste("CN", as.character(R0), sep = ""))

## Add pdb.info to site.info
site.info = cbind(site.info, pdb.info)
```

[//]: Prepare experimental data

```{r read_experimental_data}
# read experimental data
dri2.exp = read.csv(dri2.exp.fname, header = T)
site = site.info$site
dri2.exp = dri2.exp[, site]
rownames(dri2.exp) = protein.exp

Pn.exp = read.csv(Pn.exp.fname, header = T)
rownames(Pn.exp) = protein.exp

energy.exp = read.csv(energy.exp.fname, header = T)
rownames(energy.exp) = protein.exp

```

[//]: Prepare data for no selection (mut)

```{r read_and_set_data_mut}
# read simulated data

dri2.mut = read.csv(dri2.mut.fname, header = T)
site = site.info$site
dri2.mut = dri2.mut[, site]
rownames(dri2.mut) = paste("mut", rownames(dri2.mut), sep = "")

Pn.mut = read.csv(Pn.mut.fname, header = T)
rownames(Pn.mut) = paste("mut", rownames(Pn.mut), sep = "")

energy.mut = read.csv(energy.mut.fname, header = T)
rownames(energy.mut) = paste("mut", rownames(energy.mut), sep = "")
```

[//]: Prepare data for strong selection

```{r read_and_set_data_strong}
# read simulated data

dri2.strong = read.csv(dri2.strong.fname, header = T)
site = site.info$site
dri2.strong = dri2.strong[, site]
rownames(dri2.strong) = paste("strong", rownames(dri2.strong), sep = "")

Pn.strong = read.csv(Pn.strong.fname, header = T)
rownames(Pn.strong) = paste("strong", rownames(Pn.strong), sep = "")

energy.strong = read.csv(energy.strong.fname, header = T)
rownames(energy.strong) = paste("strong", rownames(energy.strong), sep = "")
```


# Structural divergence along sites

Here we study how structural divergence varies from site to site. We will compare simulated data and empirical data. 


```{r prepare-site-data}
# Prepare my data.frame
# rows of dri2 are proteins, columns are sites

tdri2.strong = as.data.frame(t(as.matrix(dri2.strong)))
tdri2.strong = cbind(site.info, tdri2.strong)
siteComparison.strong = melt(tdri2.strong, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

tdri2.mut = as.data.frame(t(as.matrix(dri2.mut)))
tdri2.mut = cbind(site.info, tdri2.mut)
siteComparison.mut = melt(tdri2.mut, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

tdri2.exp = as.data.frame(t(as.matrix(dri2.exp)))
tdri2.exp = cbind(site.info, tdri2.exp)
siteComparison.exp = melt(tdri2.exp, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

siteComparison.all.datasets = rbind(
  data.frame("dataset" = "exp", siteComparison.exp),
  data.frame("dataset" = "mut", siteComparison.mut),
  data.frame("dataset" = "strong", siteComparison.strong))

siteComparison.all.datasets = ddply(siteComparison.all.datasets, c("dataset", "protein"), mutate,
          "n.SD" = dri2/mean(dri2, na.rm = T),
          "z.SD" = vscale(dri2),
          "RSD" = sqrt(dri2),
          "n.RSD" = RSD/mean(RSD, na.rm = T),
          "z.RSD" = vscale(RSD))
```


### Mean z.RSD vs site

To measure structural divergence we use z.RSD profiles, which are the best (most similar distributions and profiles w.r.t experimental).

The following plot compares the mean z.RSD profiles, including standard errors (with .999 confidence level for the mean)
*Warning: the effective number of data should be smaller than the actual data, especially for the experimental dataset because of several factors, including smoothing of the profiles, phylogenetic relationships, etc. We need to deal with this for quantitative analysis. Probably we'll need to use bootstrapping to calculate standard errors non-parametrically*


```{r mean-z-RSD-vs-site}
# Comparison of z.RSD profiles annotated with functional information (reported active sites).
dat = siteComparison.all.datasets
dat$dataset = revalue(dat$dataset, c("strong" = "mut+sel"))

# dat.with.means and standard errors
d = ddply(dat,c("dataset","site","AA","shell"), function(x) {
                mean.z.RSD = mean(x$z.RSD, na.rm = T)
                ndata = sum(!is.na(x$z.RSD))
                se.z.RSD = sd(x$z.RSD, na.rm = T) / sqrt(ndata)
                data.frame(mean.z.RSD, ndata, se.z.RSD)
})
#d$dataset = factor(d$dataset, levels = c("exp", "mut", "strong"))
d$label = as.character(d$site)
active.sites = d$site[d$shell == 0]

p = ggplot(d,aes(x = site, y = mean.z.RSD, col = dataset, fill = dataset)) +
    geom_line() +
    geom_ribbon(aes(ymin = mean.z.RSD - f95 * se.z.RSD, ymax = mean.z.RSD + se.z.RSD), alpha = .4, show.legend = F) +
    geom_vline(xintercept = active.sites, col = "grey", show.legend = F) +
    labs(x = "site", y = "structural divergence(mean z.RSD)") +
    coord_cartesian(ylim = c(-1.5, 2.8)) +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name=NULL, values = myColors)

# Plot separate and comparative plots
p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "none")

p.facet %+% subset(d, dataset == "exp") 
p.facet %+% subset(d, dataset == "mut")
p.facet %+% subset(d, dataset == "mut+sel") 
p.facet
#p + theme(legend.position = "bottom", legend.title = element_blank())
p + theme(legend.position = "bottom")

```

VAL107 and its neighbors (including site 32) are more structurally conserved than predicted.

### Correlation between site profiles

```{r}
dat = siteComparison.all.datasets

# dat.with.means and standard errors
dat = ddply(dat,c("dataset", "site", "AA", "shell"), function(x) {
                mean.z.RSD = mean(x$z.RSD, na.rm = T)
                ndata = sum(!is.na(x$z.RSD))
                se.z.RSD = sd(x$z.RSD, na.rm = T) / sqrt(ndata)
                data.frame(mean.z.RSD, ndata, se.z.RSD)
})

dat.exp = subset(dat, dataset == "exp")
dat.mut = subset(dat, dataset == "mut")
dat.strong = subset(dat, dataset == "strong")

ggplot(dat, aes(x = site, y = mean.z.RSD, col = dataset)) + geom_line() +
    geom_text(data = subset(dat, dataset == "exp" & mean.z.RSD < -.5), aes(x = site, y = mean.z.RSD, label = site), nudge_y = -.1, size = 2)

d = dat.exp
d$d.exp.mut = dat.exp$mean.z.RSD - dat.mut$mean.z.RSD
d$d.exp.strong = dat.exp$mean.z.RSD - dat.strong$mean.z.RSD
d$d.strong.mut = dat.strong$mean.z.RSD - dat.mut$mean.z.RSD
dd = melt(data = d[, c("site", "se.z.RSD", "d.exp.mut", "d.strong.mut")], c("site", "se.z.RSD"))

ggplot(dd, aes(x = site, y = value, col = variable)) +
    geom_line() 
r = round(cor(d$d.strong.mut, d$d.exp.mut), 2)

ggplot(d, aes(x = scale(d.strong.mut), y = scale(d.exp.mut))) + 
    geom_point(aes(col = shell == 0)) +
    geom_smooth(method = "lm") +
    geom_label(aes(x = 0, y = 3, label = paste("R =", as.character(r)))) +
    labs(x = "predicted selection vs mutation (RSD change)",y="empirical vs mutation (RSD change)")
    
r.exp.mut = cor(dat.exp$mean.z.RSD, dat.mut$mean.z.RSD)
r.exp.strong = cor(dat.exp$mean.z.RSD, dat.strong$mean.z.RSD)
r.mut.strong = cor(dat.mut$mean.z.RSD, dat.strong$mean.z.RSD)
cor.df = data.frame("comparison" = c("exp vs mut", "exp vs strong", "mut vs strong"),
                    "R" = c(r.exp.mut, r.exp.strong, r.mut.strong))
knitr::kable(cor.df, digits = 2)
write.csv(dat, "structure-comparison-profiles.csv")

```

# Structural divergence along normal modes
In this section we consider structural divergence using normal mode coordinates.
Pn is the square divergence along a normal mode relative to the mean.

[//]: Prepare mode data

```{r prepare-mode-data}
mode = seq(ncol(Pn.strong))

# Prepare modeComparison.strong
tPn.strong = as.data.frame(t(as.matrix(Pn.strong)))
tPn.strong = cbind(data.frame("mode" = mode), tPn.strong)
Pn.strong.long = melt(tPn.strong, "mode", variable.name = "protein", value.name = "Pn")

tenergy.strong = as.data.frame(t(as.matrix(energy.strong)))
tenergy.strong = cbind(data.frame("mode" = mode), tenergy.strong)
energy.strong.long = melt(tenergy.strong, "mode", variable.name = "protein", value.name = "energy")
energy.strong.long$sn2 = 1/energy.strong.long$energy

modeComparison.strong = merge(energy.strong.long, Pn.strong.long, by = c("mode", "protein"))

modeComparison.strong = ddply(modeComparison.strong, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# Prepare modeComparison.mut
tPn.mut = as.data.frame(t(as.matrix(Pn.mut)))
tPn.mut = cbind(data.frame("mode" = mode), tPn.mut)
Pn.mut.long = melt(tPn.mut, "mode", variable.name = "protein", value.name = "Pn")

tenergy.mut = as.data.frame(t(as.matrix(energy.mut)))
tenergy.mut = cbind(data.frame("mode" = mode), tenergy.mut)
energy.mut.long = melt(tenergy.mut, "mode", variable.name = "protein", value.name = "energy")
energy.mut.long$sn2 = 1/energy.mut.long$energy

modeComparison.mut = merge(energy.mut.long, Pn.mut.long, by = c("mode", "protein"))
modeComparison.mut = ddply(modeComparison.mut, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# Prepare modeComparison.exp
tPn.exp = as.data.frame(t(as.matrix(Pn.exp)))
tPn.exp = cbind(data.frame("mode" = mode), tPn.exp)
Pn.exp.long = melt(tPn.exp, "mode", variable.name = "protein", value.name = "Pn")

tenergy.exp = as.data.frame(t(as.matrix(energy.exp)))
tenergy.exp = cbind(data.frame("mode" = mode), tenergy.exp)
energy.exp.long = melt(tenergy.exp, "mode", variable.name = "protein", value.name = "energy")
energy.exp.long$sn2 = 1/energy.exp.long$energy

modeComparison.exp = merge(energy.exp.long,Pn.exp.long,by=c("mode", "protein"))

# renormalize Pn
modeComparison.exp = ddply(modeComparison.exp, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# modeComparison.exp has NA's because for proteins with gaps when using Keff there're less modes
#print("Warning: modeComparison.exp <- na.omit(modeComparison.exp)")
modeComparison.exp <- na.omit(modeComparison.exp)
modeComparison.strong <- na.omit(modeComparison.strong)
modeComparison.mut <- na.omit(modeComparison.mut)

```

[//]: Add quantile regression data

```{r}
## Quantile regression 

# fit median and quantiles to simulated data and add them to data.frames
fit.strong.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.strong, method = "fn")
fit.strong.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.strong, method = "fn")
fit.strong.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.strong, method = "fn")
modeComparison.strong$rq.5 = fit.strong.rq.5$fitted.values
modeComparison.strong$rq.05 = fit.strong.rq.05$fitted.values
modeComparison.strong$rq.95 = fit.strong.rq.95$fitted.values

# fit median and quantiles to simulated data and add them to data.frames
fit.mut.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.mut, method = "fn")
fit.mut.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.mut, method = "fn")
fit.mut.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.mut, method = "fn")
modeComparison.mut$rq.5 = fit.mut.rq.5$fitted.values
modeComparison.mut$rq.05 = fit.mut.rq.05$fitted.values
modeComparison.mut$rq.95 = fit.mut.rq.95$fitted.values

# fit median and quantiles to experimental data
fit.exp.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.exp, method = "fn")
fit.exp.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.exp, method = "fn")
fit.exp.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.exp, method = "fn")
modeComparison.exp$rq.5 = fit.exp.rq.5$fitted.values
modeComparison.exp$rq.05 = fit.exp.rq.05$fitted.values
modeComparison.exp$rq.95 = fit.exp.rq.95$fitted.values

# put simulated and experimental data together

modeComparison.all.datasets = rbind(data.frame("dataset" = "exp", modeComparison.exp),
                                    data.frame("dataset" = "mut", modeComparison.mut),
                                    data.frame("dataset" = "strong", modeComparison.strong))

modeComparison.all.datasets$dataset = factor(modeComparison.all.datasets$dataset, levels = c("exp", "mut", "strong"))
```

### Pn vs. mode energy

Note: lines are the median, 0.05 and 0.95 quantiles fit using rq(Pn ~ sn2).
sn2 = 1/energy is the expected MSD of along mode n. 
The mean value of Pn should be a linear function of sn2 (Pn is normalized, sn2 is not).

Here we show the square deviation along each normal mode (normalized so that <Pn> = 1) vs. the normal-mode's energy. For the empirical data, we use the actual energy obtained from Keff for the given comparison. It is this energy the one that should be used, since it's inverse is the expected value of Pn. Lines are the .5 (median), .05, and .95 quantiles obtained using `rq()`
- Upper panel: simulated vs empirical data, shown separately.
- Lower panel: simulated vs empirical overlaped and zoomed to better see the pattern.

```{r Pn-vs-mode-energy,cache = 2}
# Plot Pn vs energy
dat = modeComparison.all.datasets
dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# p is just the median and quantiles
p.points.rq = ggplot(dat, aes(x = energy, y = Pn, col = dataset)) + 
    geom_point(alpha = .4, size = .2) +
    geom_line(aes(y = rq.5)) +
    geom_line(aes(y = rq.05)) +
    geom_line(aes(y = rq.95)) 
p = p.points.rq +  
    labs(x = "mode energy", y = "structural divergence") +
    coord_cartesian(xlim = c(0,2.5)) +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name=NULL, values = myColors)

p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "none")
p.facet %+% subset(dat, dataset == "exp")
p.facet %+% subset(dat, dataset == "mut")
p.facet %+% subset(dat, dataset == "mut+sel")
p.facet
```

### median(Pn) vs mode energy

```{r median-Pn-vs-mode-energy, cache = 2}
dat = modeComparison.all.datasets
dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# calculate dataset of medians

dat = ddply(dat,c("dataset", "mode"), function(x) {
    energy.mean = mean(x$energy)
    Pn.median = median(x$Pn)
    Pn.median.ci = sort(x$Pn)[qbinom(c(.025, .975), length(x$Pn), 0.5)]
    Pn.median.min = Pn.median.ci[1]
    Pn.median.max = Pn.median.ci[2]
    rq.5 = mean(x$rq.5)
    data.frame(energy.mean, Pn.median.min, Pn.median.max, Pn.median, rq.5)
})

p = ggplot(dat,aes(x=energy.mean,y=Pn.median,col=dataset,fill=dataset)) +
    geom_pointrange(aes(ymin=Pn.median.min,ymax = Pn.median.max),size=.2,alpha=.4) +
    geom_line(aes(y=rq.5))  +
    labs(x="mode energy",y="structural divergence (median Pn) ") +
    scale_colour_manual(name = NULL,values = myColors) + 
    scale_fill_manual(name=NULL,values=myColors)+
    coord_cartesian(xlim=c(0,1)) 


p.facet = p + facet_grid(dataset~.,switch="y") + theme(legend.position="none")
p.facet %+% subset(dat,dataset =="exp")
p.facet %+% subset(dat,dataset == "mut")
p.facet %+% subset(dat,dataset == "mut+sel")
p.facet
p + theme(legend.position = "bottom")
    

```

### Correlation between mode profiles

```{r}
dat = modeComparison.all.datasets
#dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# calculate dataset of medians
dat = ddply(dat, c("dataset", "mode"), function(x) {
    energy.mean = mean(x$energy)
    Pn.median = median(x$Pn)
    data.frame(energy.mean, Pn.median)
})

dat = subset(dat, mode < 101)
dat.exp = subset(dat, dataset == "exp")
dat.mut = subset(dat,dataset == "mut")
dat.strong = subset(dat,dataset == "strong")

r.exp.mut = cor(dat.exp$Pn.median, dat.mut$Pn.median)
r.exp.strong = cor(dat.exp$Pn.median, dat.strong$Pn.median)
r.mut.strong = cor(dat.mut$Pn.median, dat.strong$Pn.median)
cor.df = data.frame("comparison" = c("exp vs mut", "exp vs strong", "mut vs strong"),
                    "R" = c(r.exp.mut, r.exp.strong, r.mut.strong))
knitr::kable(cor.df, digits = 2)

```

[//]: Write output for further analysis with other programs

```{r write-data-to-output}
write.csv(siteComparison.exp, "site-structure-comparison-exp.csv",row.names=F)
write.csv(modeComparison.exp, "mode-structure-comparison-exp.csv",row.names=F)
```

